Evolution of \\ ice sheet geometry \\ using Stokes dynamics

### 2021-06-22 talk/talk.tex
Simplifying edits after practice run (23 min).

\begin{frame}{overview}
\Large
\begin{enumerate}
\item explain the problem I'm obsessed with
\item discuss the current barriers to success
\end{enumerate}

\normalsize
\vspace{10mm}
\begin{itemize}
\item<2-3> work in progress
\item<2-3> I am giving this talk at 4am my time
    \begin{itemize}
    \item<3> low expectations, please
    \end{itemize}
\end{itemize}
\end{frame}

@@

\vspace{-1mm}
\begin{itemize}
\item glacier geometry evolution according to the climate?
    \begin{itemize}
    \item simplest such problem (grounded, nonsliding, isothermal)
    \end{itemize}
\item assume time-independent data on fixed $\Omega \subset \RR^2$:
    \begin{itemize}
    \item (net) climatic mass balance $a(x,y)$ \hfill $\gets$ \emph{$a>0$ for accumulation}
    \item bed elevation $b(x,y)$
    \end{itemize}
\item \emph{to compute}: surface elevation $s(t,x,y)$ on $\Omega$ \emph{and} icy domain $\Lambda_s \subset \RR^3$
\end{itemize}
\end{frame}

@@

    \begin{itemize}
    \item $\bn_s = \left<-s_x,-s_y,1\right>$  \, is an upward normal to the ice surface
    \item where \emph{not} on the ice: $a\le 0$
    \end{itemize}
\end{itemize}

\bigskip
\begin{columns}
    \column{0.55\textwidth}
        \begin{itemize}
        \item assumed for this talk:
            \begin{itemize}
            \item $s$ well-defined (no overhangs)
            \item $s=b$ where ice is not present

                \begin{itemize}
                \item[{\color{black} $\circ$}] $s$ is defined on all of $\Omega$
                \end{itemize}
            \end{itemize}
        \end{itemize}
    \column{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{figs/simpledomain.png}
\end{columns}
\end{frame}

@@

    \begin{itemize}
    \item very similar to steady IGP
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item almost no one is solving such a implicit or steady IGP
    \begin{itemize}
    \item except Bueler (2016), Brinkerhoff et al (2017) (\emph{shallow})
    \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, fake ice layer})
    \item none are scalable (\emph{single level})
    \end{itemize}
\item what are people doing instead, for production science?
    \begin{itemize}
    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
    \item often with shallow approximations, e.g.~PISM run below
    \end{itemize}

@@

\begin{itemize}
\item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0,$$
extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
\item regarding this Stokes problem (for \emph{fixed} geometry):
    \begin{itemize}
    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz 2011)
    \item optimal solver exists (Isaac et al 2015)
    \end{itemize}
\item $\therefore$ \, $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
    \begin{itemize}
    \item which is a regularity conjecture
    \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item using $\Phi$ gives a cleaner NCP over $\Omega$ for steady IGP:
\begin{align*}
s - b &\ge 0 \\
\Phi(s) - a &\ge 0 \\
(s - b) (\Phi(s) - a) &= 0
\end{align*}

\item NCP (strong form) $\leftrightarrow$ VI (weak form)

\begin{block}{steady IGP weak form}
find admissible $s \in \mathcal{K} = \{r \ge b\} \subset W^{1,\qq}(\Omega)$ so that
    $$\ip{\Phi(s)-a}{r-s} \ge 0$$
for all $r \in \mathcal{K}$
\end{block}

@@

\begin{itemize}
\small
\item $\Phi : \mathcal{K} \subset W^{1,\qq}(\Omega) \to W^{-1,\pp}(\Omega)$ where $\qq=4$ and $\pp=4/3$
\item the Stokes stress balance is nonlocal so $\Phi$ is ``dense''
    \begin{itemize}
    \item $\Phi$ is some kind of integral operator (\emph{convolve with Stokeslets})
    \item $\Phi$ is short range (\emph{concentrated near the diagonal})
    \end{itemize}
\item $\Phi \approx \Phi_{\text{SIA}}$ for shallow glaciers, a differential operator:
    $$\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right]$$

@@

\item this strategy gives an $O(m\log m)$ solver for the Laplacian obstacle problem (Tai, 2003)
    \begin{itemize}
    \item \alert{a multilevel VI strategy is the only hope for (near) optimality}
    \end{itemize}
\item for IGP:
    \begin{itemize}
    \item for nonlinear VI, need additional FAS strategy {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
    \item each residual is nonlocal and expensive {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
    \item<2> {\color{ForestGreen} strategy implemented}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item now all the evil is in the smoother
\item smoother options from weaker to stronger:
    \begin{itemize}
    \item nonlinear Richardson?
        \begin{itemize}
        \item[{\color{black} $\circ$}] not promising because degenerate
        \end{itemize}
    \item nonlinear pointwise smoothers (GS, Jacobi)?
        \begin{itemize}
        \item[{\color{black} $\circ$}] \only<1>{GS extremely expensive because residual nonlocal} \only<2>{\alert{GS extremely expensive because residual nonlocal}}
        \item[{\color{black} $\circ$}] seem to fail
        \end{itemize}
    \item exploit SIA analogy?
        \begin{itemize}
        \item[{\color{black} $\circ$}] pointwise smoothers work for SIA
        \end{itemize}
    \item nonlinear and nonlocal \texttt{PCPatch} smoothers?
        \begin{itemize}
        \item[{\color{black} $\circ$}] what does this even mean?
        \end{itemize}
    \item reduced-space Newton using banded Jacobian approximation?
        \begin{itemize}
        \item[{\color{black} $\circ$}] finite-difference Jacobian entries using pseudo-coloring
        \end{itemize}
    \end{itemize}

\medskip
\item<2> status report:
    \begin{itemize}
    \item \alert{nothing works well so far}
    \item NewtonRS and SIA-based smoothers most promising?
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item \emph{GOAL 1}:\, no shallow approximations in flow physics {\Large \strut} \only<2-3>{{\Large \color{ForestGreen} $\checkmark$}}
    \begin{itemize}
    \item<2-3> my Stokes solver is working in Firedrake on extruded mesh
    \item<2-3> solver choices by (Isaac et al, 2015) putatively optimal
    \end{itemize}
\item \emph{GOAL 2}:\, beat explicit time-stepping and enable long-time paleo simulations {\Large \strut} \only<3>{\Large \alert{:-(}}
    \begin{itemize}
    \item<2-3> problem is
        \begin{itemize}
        \item<2-3>[{\color{black} $\circ$}] constrained (nonlinear VI on $\mathcal{K} = \{s\ge b\}$)
        \item<2-3>[{\color{black} $\circ$}] diffusive in the large (Stokes $\approx$ SIA)
        \item<2-3>[{\color{black} $\circ$}] nonlocal (Stokes)
        \end{itemize}
    \item<2-3> {\color{ForestGreen} MCD FAS path to optimality for solving the VI}
    \item<3> \alert{stuck on finding a robust optimal smoother}
    \end{itemize}

\normalsize
\bigskip\bigskip
\item<3> \emph{thanks for your attention \dots questions?}
\end{itemize}
\end{frame}

>>>>

\begin{frame}{overview}
\Large
\begin{enumerate}
\item explain the problem I'm working on
\item mention the (current) barriers to success
\end{enumerate}

\normalsize
\vspace{10mm}
\begin{itemize}
\item work in progress
\item I am giving this talk at 4am my time
    \begin{itemize}
    \item low expectations, please
    \end{itemize}
\end{itemize}
\end{frame}

@@

\vspace{-1mm}
\begin{itemize}
\item how does glacier geometry evolve in response to the climate?
\item only the simplest version (grounded, nonsliding, isothermal)
\item assume time-independent data on fixed $\Omega \subset \RR^2$:
    \begin{itemize}
    \item (net) climatic mass balance $a(x,y)$ \hfill $\gets$ \emph{$a>0$ for accumulation}
    \item bed elevation $b(x,y)$
    \end{itemize}
\item \emph{to compute}: surface elevation $s(t,x,y)$ on $\Omega$

\hspace{20.5mm} icy domain $\Lambda_s \subset \RR^3$
\end{itemize}
\end{frame}

@@

    \begin{itemize}
    \item $\bn_s = \left<-s_x,-s_y,1\right>$  \, is an upward normal to the ice surface

\vspace{4mm}
    \item where \emph{not} on the ice: $a\le 0$
    \end{itemize}
\end{itemize}

\bigskip
\begin{columns}
    \column{0.55\textwidth}
        \begin{itemize}
        \item assumed:
            \begin{itemize}
            \item $s$ well-defined (no overhangs)
            \item $s=b$ where ice is not present

                \begin{itemize}
                \item[{\color{black} $\therefore$}] $s$ is defined on all of $\Omega$
                \end{itemize}
            \end{itemize}
        \end{itemize}
    \column{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{figs/simpledomain.png}
\end{columns}
\end{frame}

@@

    \begin{itemize}
    \item \alert{very similar to steady IGP}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item almost no one is solving such a implicit or steady IGP
    \begin{itemize}
    \item except Bueler (2016), Brinkerhoff et al (2017) (\emph{shallow})
    \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, fake ice layer})
    \item none are scalable (\emph{single level})
    \end{itemize}
\item what are people doing instead, for production science?
    \begin{itemize}
    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
    \item usually shallow approximations, e.g.~PISM run below
    \end{itemize}

@@

\begin{itemize}
\item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0,$$
extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
\item regarding this Stokes problem (for \emph{fixed} geometry):
    \begin{itemize}
    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz 2011)
    \item optimal solver exists (Isaac et al 2015)
    \end{itemize}
\item[$\therefore$] $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
    \begin{itemize}
    \item which is a regularity conjecture
    \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item using $\Phi$ gives a cleaner NCP over $\Omega$ for the steady IGP:
\begin{align*}
s - b &\ge 0 \\
\Phi(s) - a &\ge 0 \\
(s - b) (\Phi(s) - a) &= 0
\end{align*}

    \begin{itemize}
    \item recall NCP (strong form) $\leftrightarrow$ VI (weak form)
    \end{itemize}

\begin{block}{steady IGP weak form}
find admissible $s \in \mathcal{K} = \{r \ge b\} \subset W^{1,\qq}(\Omega)$ so that
    $$\ip{\Phi(s)-a}{r-s} \ge 0 \qquad \text{ for all } r \in \mathcal{K}$$
\end{block}

@@

\begin{itemize}
\small
\item $\Phi : \mathcal{K} \subset W^{1,\qq}(\Omega) \to W^{-1,\pp}(\Omega)$ where $\qq=4$ and $\pp=4/3$
\item the Stokes stress balance is nonlocal
    \begin{itemize}
    \item $\Phi$ is integral operator (\emph{convolve with Stokeslets})
    \item $\Phi$ is short range (\emph{dense but concentrated near the diagonal})
    \end{itemize}
\item $\Phi \approx \Phi_{\text{SIA}}$ for shallow glaciers, a differential operator:
    $$\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right]$$

@@

\item this strategy gives an $O(m\log m)$ solver for the Laplacian obstacle problem (Tai, 2003)
    \begin{itemize}
    \item \alert{a multilevel VI strategy is the only hope for (near) optimality}
    \end{itemize}
\item for IGP:
    \begin{itemize}
    \item for nonlinear VI, need additional FAS strategy {\Large\strut} {\color{ForestGreen} \, {\Large $\checkmark$}}
    \item \dots but each residual is nonlocal and expensive
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item now all the evil is in the smoother
\item smoother options from weaker to stronger:
    \begin{itemize}
    \item nonlinear Richardson?
    \item exploit SIA analogy?
        \begin{itemize}
        \item[{\color{black} $\circ$}] pointwise smoothers work for SIA
        \end{itemize}
    \item nonlinear pointwise smoothers (GS, Jacobi)?
        \begin{itemize}
        \item[{\color{black} $\circ$}] \alert{GS extremely expensive because residual nonlocal}
        \item[{\color{black} $\circ$}] seem to fail
        \end{itemize}
    %\item nonlinear and nonlocal \texttt{PCPatch} smoothers?
    %    \begin{itemize}
    %    \item[{\color{black} $\circ$}] what does this even mean?
    %    \end{itemize}
    \item reduced-space Newton using banded Jacobian approximation?
        \begin{itemize}
        \item[{\color{black} $\circ$}] finite-difference Jacobian entries using pseudo-coloring
        \end{itemize}
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{status report}

\begin{itemize}
\item \alert{nothing works well so far}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item \emph{GOAL 1}:\, no shallow approximations in flow physics
    \begin{itemize}
    \item[{\Large \color{ForestGreen} $\checkmark$}] Stokes solver working in Firedrake on extruded mesh
    \item[{\Large \color{ForestGreen} $\checkmark$}] solver choices by (Isaac et al, 2015) putatively optimal
    \end{itemize}
\item \emph{GOAL 2}:\, beat explicit time-stepping
    \begin{itemize}
    \item[{\Large \color{ForestGreen} $\checkmark$}] observe that problem is
        \begin{itemize}
        \item[{\color{black} $\circ$}] constrained (nonlinear VI on $\mathcal{K} = \{s\ge b\}$)
        \item[{\color{black} $\circ$}] diffusive in the large (Stokes $\approx$ SIA)
        \item[{\color{black} $\circ$}] nonlocal (Stokes)
        \end{itemize}
    \item[{\Large \color{ForestGreen} $\checkmark$}] MCD FAS path to optimality for solving the VI
    \item[{\Large \alert{:-(}}] \alert{stuck on finding a robust, efficient smoother}
    \end{itemize}
\end{itemize}

\begin{block}{ultimate goal}
for better science, enable long-time (paleo) simulations without shallow approximations
\end{block}

\emph{thanks for your attention \dots questions?}
\end{frame}
